\chapter{Implementation}

At this point, we have arrived at a suitable architecture and design for \texttt{JHygge}, the new Hygge compiler, so this chapter concerns the actual
implementation. Here, we emphasize the use of the new functional features of modern Java (JDK 21 and up) and present the highlights
of the final implementation of \texttt{JHygge}. 

\section{Use of Functional Java: Records, Sealed Interfaces, Pattern matching and Monads}

As the purpose of this thesis is also to explore the newer functional features of \texttt{Java}. The existing implementation of \texttt{hyggec}
is indeed very functional, so we attempted to write the implementations of both the AST, typechecker and interpreter in such a way that
the \texttt{JHygge} versions are as close to those of \texttt{hyggec}, disregarding the architectural and design differences. Some of
the newer functional features of \texttt{Java} we wanted to investigate include records (immutable classes), sealed interfaces (discriminated unions),
switch expressions (pattern matching) and the introduction of some monads, most notably \texttt{Optional}. In the following sections,
we show concrete examples of implementation, where we have attempted to use said features and compare them to the corresponding code
in \texttt{hyggec}.

\subsection{Records and Sealed Interfaces for functional data modelling}

The \texttt{Node} and \texttt{Expr} data model for the AST from \texttt{hyggec} is one of the parts that we wish to carry over in \texttt{JHygge}.
This requires support for discriminated unions and recursive data type, the former of which was introduced in JDK 17 in the form of sealed
interfaces; one declares a finite number of exact specializations upon defining the interface itself. This gives the \texttt{Java} compiler
information about the possible specialization, which it can use to perform case analysis and thus enforce exhaustive pattern matching. Let's
see records and sealed interfaces in action in the form of \texttt{JHygge}s implementation of \texttt{Expr} in figure \ref{fig:expr_sealed_interface}.

\begin{figure}[H]
\centering 
\begin{lstlisting}[language=Java]
// From Expr.java, reformatted to be more compact and fit within the margins
public sealed interface Expr<E, T>
    permits Expr.UnitVal, Expr.BoolVal, Expr.IntVal, Expr.FloatVal,
        Expr.StringVal, Expr.Var, Expr.BinaryOperator, Expr.Not,
        Expr.ReadInt, Expr.ReadFloat, Expr.Print, Expr.PrintLn,
        Expr.If, Expr.Seq, Expr.Type, Expr.Ascription,
        Expr.Assertion, Expr.Let, Expr.Assign, Expr.While,
        Expr.Lambda, Expr.Application, Expr.StructCons,
        Expr.FieldSelect, Expr.Pointer, Expr.UnionCons,
        Expr.Match {
    // ...
}
\end{lstlisting}
\caption{The implementation of Hygge expressions as a \texttt{sealed interface} specifying the exact allowed specializations.}
\label{fig:expr_sealed_interface}
\end{figure}

Similar to discriminated unions in \texttt{F\#}, all cases are declared for the sealed interface \texttt{Expr}. However, unlike \texttt{F\#},
the cases have to be defined as separate classes, records or interfaces. In the case that a specialization is a class, they have to be \texttt{final}.
Let us examine an example of one of the mutually-recursive specializations of \texttt{Expr}, which also happens to be a \texttt{record}, namely the \texttt{Not}-expression for which the code is shown in figure \ref{fig:not_record}.

\begin{figure}[H]
\centering
\begin{lstlisting}[language=Java]
  // From Expr.java, Not<E, T> is an inner class of Expr<E, T>
  record Not<E, T>(Node<E, T> arg) implements Expr<E, T> {
    @Override
    public String prettyPrint(int indentation) {
      indentation += 2;
      return "Not(arg=" + arg.expr().prettyPrint(indentation) + ")";
    }
  }
\end{lstlisting}
\caption{An implementation of the Hygge \texttt{Not} expression in the form of a \texttt{record}.}
\label{fig:not_record}
\end{figure}

Declaring \texttt{Not<E, T>} as a \texttt{record} has the same effect as if it had been declared as a class with private \texttt{final}
attributes that inherit from the \texttt{java.lang.Record}, where each attribute only has a getter, but not a setter, along with implicit declarations
of some utility methods, such as the \texttt{Object.equals()}-, \texttt{Object.hashCode()}- and \texttt{Object.toString()}-method. Now, this example uses the
\texttt{record} declaration syntax, which does all of the things mentioned above to remove boilerplate to ease the burden of the programmer.

Another important thing to note about the immutability of Java records is that they are shallowly immutable, meaning that although that the record
attributes, also called components, cannot be altered, the attributes of the record components can be mutable. That is, this \texttt{Not<E, T>} record
has an \texttt{arg} component of type \texttt{Node<E, T>}, which is also a record, but had \texttt{arg} been declared as a class with mutable
attributes, the programmer could still change the value of those attributes eventhough \texttt{Not<E, T>} is a record.

A record is also \texttt{final} in the sense that extending record types through inheritance is prohibitted; it cannot be used as a base for
further specialization. This makes sense in the context of sealed interfaces and pattern matching, where it would otherwise be difficult
to perform case analysis and reason about pattern exhaustion had some variant in the discriminated union been a regular class that could
have had an unknown number of specializations.

\subsection{Switch expressions for case-by-case definitions and dispatching}

In the preceding section, we mentioned pattern matching when discussing the use of discriminated unions. In order to handle the different variants
of a discriminated union, pattern matching gives the programmer a way to not only branch based on the value of the active variant, but also on
the properties and the structure of the value (in some languages). With JEP XXX, the \texttt{switch}-expression was introduced in Java, which behaves
similarly to the \texttt{match}-expressions in languages like \texttt{F\#} and \texttt{Rust}. Unlike the existing \texttt{switch}-statement,
the newer \texttt{switch}-expression returns a value and one can use the \texttt{when}-clause to branch to multiple different cases for the
same variant in a discriminated union. This is useful in multiple compiler phases every time the \texttt{Expr}-type is involved. Take
the \texttt{reduce}-method as an example; in order to reduce each variant of the \texttt{Expr}-type, the \texttt{switch}-expression is used
to delegate the responsibility of expression reduction to the appropriate reducer as per our design proposal. An excerpt for the code
for the \texttt{reduce}-method from \texttt{StandardHyggeInterpreter} is shown on figure \ref{fig:switch_expression_values}:

\begin{figure}[H]
\centering
\begin{lstlisting}[language=Java]
  // From StandardHyggeInterpreter.java, slightly reformatted
  @Override
  public Optional<InterpreterResult<E, T>> reduce(RuntimeEnvironment<E, T> env, Node<E, T> node) {
    return switch (node.expr()) {
      case UnitVal<E, T> _,
          BoolVal<E, T> _,
          IntVal<E, T> _,
          FloatVal<E, T> _,
          StringVal<E, T> _,
          Pointer<E, T> _ ->
          Optional.empty();
      case Var<E, T> v when env.isMutable(v.name()) ->
          InterpreterResult.of(env, env.mutables().get(v.name()));
      case Var<E, T> _ -> Optional.empty();
      case BinaryOperator<E, T> v -> binaryOperatorReducer.reduce(this, env, node, v);
      case Not<E, T> v ->
          (v.arg().expr() instanceof BoolVal<E, T> b)
              ? InterpreterResult.of(env, node.with(b.not()))
              : reduce(env, v.arg())
                  .map(a -> new InterpreterResult<>(env, node.with(new Not<>(a.node()))));
      // Remaining cases redacted for conciseness...
    };
  }
\end{lstlisting}
\caption{A simple use of a \texttt{switch}-expression to branch based on the Hygge expression contained in a \texttt{Node}. Undiscussed cases have been removed for the sake of brevity.}
\label{fig:switch_expression_values}
\end{figure}

As we can see in the listing, the cases can group multiple variants to branch to the same single expression, as with
the value variants \texttt{UnitVal<E, T>}, etc., which is implicitly returned/yielded. For variants like \texttt{Var<E, T>}, there are
two different cases to handle, where the \texttt{when}-clause is used to disguish between these cases without having to resort to
additional constructs for control flow creating unwanted nesting. One might point out, that this essentially a functional programming
version of the Visitor pattern known from Object-oriented design. So, this is a case, where both FP and OOP style want to solve the
same problem, but do it in different yet still somewhat similar ways.

However, while pattern matching with \texttt{switch}-expressions can be useful, there is still limited support for it as one
can only pattern match on reference types (classes, interfaces, enums, etc.) as of Java 24 (a proposal for use with primitives is in preview with JEP XXX).
That is, one cannot easily do pattern matching on the structure or value of a type. Take the list data structure as an example;
in \texttt{F\#}, one can match not only against the type, but also the value and patterns of values, i.e. matching against \texttt{x :: xs}
would mean that there is at least one element in the list and maybe more. These kinds of patterns are not supported in Java 24, so working with lists and
map types are less straight-forward. Take the \texttt{Seq}-record for the sequencing expression as an example; it has a record component,
\texttt{nodes}, which is of type \texttt{List<Node<E, T>>}. There is a case during interpretation, where the interpreter should apply
different reductions for a sequence expression depending on the number of successive expressions, namely none, one and more than one.
While one could use a \texttt{switch}-expression with \texttt{when}-clauses, this would create deep and nested code impairing readability
due to indentation. Instead, it is simpler to branch on the size of the list using \texttt{if}-statements as shown in figure \ref{fig:seq_reduce_implementation},
which contains the \texttt{Java}-implementation of the reduction rules for the sequencing expression.

\begin{figure}[H]
\centering
\begin{lstlisting}[language=Java]
  // From StandardHyggeControlFlowReducer.java
  @Override
  public Optional<InterpreterResult<E, T>> reduce(
      HyggeInterpreter<E, T> interpreter,
      RuntimeEnvironment<E, T> env,
      Node<E, T> node,
      Seq<E, T> seqExpr) {
    var nodes = seqExpr.nodes();

    if (nodes.isEmpty()) {
      return InterpreterResult.of(env, node.with(new UnitVal<>()));
    }

    if (nodes.size() == 1) {
      var last = nodes.getFirst();
      return ASTUtility.isValue(last)
          ? InterpreterResult.of(env, node.with(last.expr()))
          : interpreter
              .reduce(env, last)
              .flatMap(
                  result -> InterpreterResult.of(result.env(), node.with(result.node().expr())));
    }

    var first = nodes.getFirst();
    var rest = nodes.subList(1, nodes.size());
    return !ASTUtility.isValue(first)
        ? interpreter
            .reduce(env, first)
            .flatMap(
                result ->
                    InterpreterResult.of(
                        result.env(),
                        node.with(
                            new Seq<>(
                                Stream.concat(Stream.of(result.node()), rest.stream()).toList()))))
        : InterpreterResult.of(env, node.with(new Seq<>(rest)));
  }
\end{lstlisting}
\caption{Implementation for the reduction rules of \texttt{Seq}-expressions without the use of a \texttt{switch}-expression}
\label{fig:seq_reduce_implementation}
\end{figure}

This implementation is not written in a style that is as functional as the \texttt{hyggec} counterpart and it seems that
this is one of the compromises of a language like \texttt{Java}, which was not designed to be functional from the beginning.
Also, notice that there is a need to create a local variable, \texttt{nodes}, to \texttt{seqExpr.nodes()} because pattern
matching isn't supported for method arguments as it is in \texttt{F\#}.

\subsection{\texttt{Optional} for monadic error handling}

The Option/Maybe and Result/Either types are ways to handle errors for effectful/impure code in a way that resembles pure code.
This is in contrast to exceptions, which is the de facto default approach to error handling in \texttt{Java}. The benefit
of using monadic error handling is that it becomes immediately obvious that a function can fail from its type signature.
Furthermore, the type also communicates the kind of errors that can occur. The Option/Maybe type indicates that the function
may produce something or nothing, and in the case of nothing, there is no information about why. This is useful in cases,
where when a failure occurs, the programmer should not be concerned about why it failed. On the other hand, the Result/Either
type is used when the programmer is interested in the cause of failure and want to act on it.

In \texttt{Java}, there is the \texttt{java.lang.Optional} type, which is an implementation of the Option/Maybe monad. It
supports the $>>=$ (pronounced: ``bind'') and \texttt{return} operations in the form of the \texttt{.flatMap()}-method
and \texttt{Optional.of()}, respectively. In \texttt{JHygge}, this has been used for the implementation of the interpreter
reduction rules based on the algorithm from \texttt{hyggec} as reduction is applied until it is impossible to reduce further.
At that point, the last successful reduced \texttt{Node} is returned. In figure \ref{fig:struct_reducer_monads}, we show an excerpt from
\texttt{StandardStructReducer.java}, which showcases the use of \texttt{Optional} in the \texttt{reduce()}-method.

\begin{figure}[H]
\centering  
\begin{lstlisting}[language=Java]
// From StandardStructReducer.java, all other methods have been redacted ...
public class StandardStructReducer<E, T> implements StructReducer<E, T> {

  @Override
  public Optional<InterpreterResult<E, T>> reduce(
      HyggeInterpreter<E, T> interpreter,
      RuntimeEnvironment<E, T> env,
      Node<E, T> node,
      FieldSelect<E, T> fieldSelectExpr) {

    if (fieldSelectExpr.target().expr() instanceof Expr.Pointer<E, T>(int address))
      return FunctionalUtilities.tryFind(env.ptrInfo(), address)
          .flatMap(
              fields -> {
                var offset = fields.indexOf(fieldSelectExpr.field());
                return InterpreterResult.of(env, env.heap().get(address + offset));
              });

    if (!ASTUtility.isValue(fieldSelectExpr.target()))
      return interpreter
          .reduce(env, fieldSelectExpr.target())
          .map(result -> result.with(new FieldSelect<>(result.node(), fieldSelectExpr.field())));

    return Optional.empty();
  }

}
\end{lstlisting}
\caption{Implementation of the reduction rules for Hygge \texttt{struct}s making use of the \texttt{Optional} monad.}
\label{fig:struct_reducer_monads}
\end{figure}

As shown in the listing above, the both the \texttt{.flatMap()} and \texttt{.map()} operations as well as \texttt{Optional.of()}
indirectly (this is hidden in \texttt{InterpreterResult.of()} and \texttt{FunctionalUtilities.tryFind()}). Notice how using
this technique emphasizes the ``happy path'' of the program as error propagation becomes implicit due to the functor and monad
operations \texttt{.map()} and \texttt{.flatMap()} allow the programmer to focus on the data transformations instead of error
handling.

\section{Parsing Hygge source files with ANTLR4}

As mentioned in the chapter on Architecture and Design, we have selected ANTLR4 as the parser generator of choice. For starters,
we have translated the \texttt{FsLexYacc} lexer and parser files, \texttt{Lexer.fsl} and \texttt{Parser.fsy}, to ANTLR4
\texttt{.g4}-format. In figure \ref{fig:hygge_grammar_antlr4}, we show the complete \texttt{.g4} grammar file for the final version of \texttt{JHygge}.

\begin{figure}[H]
\centering  
\begin{lstlisting}
grammar Hygge;

@header {
package io.github.hacktheoxidation.hygge;
}

prog: expression EOF;

expression: TYPE ID EQUAL pretype SEQ expression                                        # TypeAlias
    | LET (MUTABLE)? variable (ASCRIPTION pretype)? EQUAL simpleExpr SEQ expression     # LetDeclare
    | FUN variable parenArgTypesSeq ASCRIPTION pretype EQUAL simpleExpr SEQ expression  # FunDeclare
    | curlyExpr expression                                                              # Curly
    | sequenceExpr                                                                      # Seq
    ;

curlyExpr: LCURLY expression RCURLY                                       # Scope
    ;

sequenceExpr: simpleExpr SEQ expression                                   # Sequence
    | simpleExpr                                                          # Simple
    ;

simpleExpr: ifExpr                                                        # If
    | curlyExpr                                                           # Brackets
    | WHILE simpleExpr DO simpleExpr                                      # While
    | primaryExpr LARROW simpleExpr                                       # Assignment
    | FUN parenArgTypesSeq RARROW simpleExpr                              # Lambda
    | MATCH simpleExpr WITH LCURLY matchCases RCURLY                      # Match
    ;

ifExpr: IF simpleExpr THEN simpleExpr ELSE simpleExpr                     # IfThenElse
    | orExpr                                                              # Or
    ;

orExpr: orExpr OR andExpr                                                 # LogicalOr
    | andExpr                                                             # And
    ;

andExpr: andExpr AND relExpr                                              # LogicalAnd
    | relExpr                                                             # Rel
    ;

relExpr: addExpr EQUAL addExpr                                            # Equality
    | addExpr LESS addExpr                                                # RelLess
    | addExpr                                                             # Add
    ;

addExpr: addExpr ADD multExpr                                             # ArithAdd
    | multExpr                                                            # Mult
    ;

multExpr: multExpr MUL unaryExpr                                          # ArithMul
    | unaryExpr                                                           # Unary
    ;

unaryExpr: NOT unaryExpr                                                  # LogicalNot
    | READINT ('(' ')' | '()')                                            # ReadInt
    | READFLOAT ('(' ')' | '()')                                          # ReadFloat
    | PRINT '(' simpleExpr ')'                                            # Print
    | PRINTLN '(' simpleExpr ')'                                          # PrintLine
    | ASSERT '(' simpleExpr ')'                                           # Assertion
    | ascriptionExpr                                                      # Ascription
    | primaryExpr parenExprCommaSeq                                       # Application
    ;

ascriptionExpr: primaryExpr ASCRIPTION pretype                            # TypeAscription
    | primaryExpr                                                         # Primary
    ;

primaryExpr: LPAREN simpleExpr RPAREN                                     # Parens
    | value                                                               # Val
    | variable                                                            # Var
    | STRUCT LCURLY structFieldInitSeq RCURLY                             # StructCons
    | primaryExpr DOT field                                               # FieldSelect
    | label LCURLY expression RCURLY                                      # UnionCons
    ;


value : INT | FLOAT | BOOL | STRING | UNIT ;
variable : ID ;
field : ID ;
label: ID ;

pretype : parenTypesSeq RARROW pretype
    | STRUCT LCURLY structFieldTypeSeq RCURLY
    | UNION LCURLY unionLabelTypeSeq RCURLY
    | ID
    ;


parenArgTypesSeq: UNIT
    | LPAREN (variable ASCRIPTION pretype (COMMA variable ASCRIPTION pretype)*) RPAREN
    ;

parenTypesSeq: UNIT
    | LPAREN (pretype (COMMA pretype)*) RPAREN
    ;

parenExprCommaSeq: UNIT
    | LPAREN (simpleExpr (COMMA simpleExpr)*) RPAREN
    ;

structFieldInitSeq: field EQUAL simpleExpr (SEQ field EQUAL simpleExpr)*;

structFieldTypeSeq: field ASCRIPTION pretype (SEQ field ASCRIPTION pretype)*;

unionLabelTypeSeq: label ASCRIPTION pretype (SEQ label ASCRIPTION pretype)*;

matchCases: label LCURLY variable RCURLY RARROW simpleExpr (SEQ label LCURLY variable RCURLY RARROW simpleExpr)*;

// Reserved keywords and operators
ASSERT : 'assert' ;
READINT : 'readInt' ;
READFLOAT : 'readFloat' ;
PRINT : 'print' ;
PRINTLN : 'println';
NOT : 'not' ;
ADD : '+' ;
SUB : '-' ;
MUL : '*' ;
LESS : '<' ;
AND : 'and' ;
OR : 'or' ;
IF : 'if' ;
THEN : 'then' ;
ELSE : 'else' ;
LET : 'let' ;
TYPE : 'type' ;
EQUAL : '=' ;
LPAREN : '(' ;
RPAREN : ')' ;
LCURLY : '{' ;
RCURLY : '}' ;
ASCRIPTION : ':' ;
SEQ : ';' ;
COMMA : ',' ;
DOT : '.' ;
LARROW : '<-' ;
RARROW : '->' ;
MUTABLE : 'mutable' ;
WHILE : 'while' ;
DO : 'do' ;
FUN : 'fun' ;
STRUCT : 'struct' ;
UNION : 'union' ;
MATCH : 'match' ;
WITH : 'with' ;

// Values and Names
INT : [0-9]+ ;
FLOAT : INT '.' INT ('f' | 'F') ;
BOOL : 'true' | 'false' ;
STRING : '"' (~["\\\r\n] | EscapeSequence)* '"' ;
UNIT : '()' ;
ID: [a-zA-Z_][a-zA-Z_0-9]* ;
WS: [ \t\n\r\f]+ -> skip ;
LINE_COMMENT: '//' ~[\r\n]* -> skip ;

fragment EscapeSequence: '\\' [btnfr"'\\] ; 
\end{lstlisting}
\caption{Complete ANTLR4 \texttt{.g4} grammar specification for the Hygge language in the final version of \texttt{JHygge}.}
\label{fig:hygge_grammar_antlr4}
\end{figure}

Initially, this grammar file was syntactically more similar to the E-BNF for the \texttt{Hygge} specification as an attempt to
simplify the grammar file as much as possible. However, the \texttt{Hygge} grammar specification does not take operator
precedence into account, so it was later decided to factor out left-recursion and enforce operator precedence by separating
production rules similar to that of \texttt{Parser.fsy} in \texttt{hyggec}.

By default, \texttt{ANTLR4} generates both a lexer and a parser along with classes for implementing listeners and visitors.
One could implement the rest of \texttt{JHygge} using the \texttt{ANTLR4} AST representation, but this AST is intended to
be consumed or transformed by listeners and visitors, respecitively. Instead, we want a more functional model of an AST.
In order to breach the gap between the \texttt{ANTLR4} representation and our desired \texttt{Node} and \texttt{Expr} data model,
we decided to create a visitor, \texttt{AntlrHyggeNodeVisitor}, that provides this transformation. In the listing below, we
see an excerpt of \texttt{AntlrHyggeNodeVisitor}, namely the \texttt{visitAssertion}, which transforms a \texttt{HyggeParser.AssertionContext}
to a \texttt{Node} with an \texttt{Expr.Assertion}-expression, as well as the code for the \texttt{Expr.Assertion} record.
This implementation is shown in figure \ref{fig:antlr_node_visitor}.

\begin{figure}[H]
\centering 
\begin{lstlisting}[language=Java]
 
  // From AntlrHyggeNodeVisitor.java
  @Override
  public Node<Object, Object> visitAssertion(HyggeParser.AssertionContext ctx) {
    return new Node<>(
        new Expr.Assertion<>(ctx.simpleExpr().accept(this)),
        Position.from(ctx.start, ctx.stop),
        null,
        null);
  }

  // From Expr.java
  record Ascription<E, T>(PreTypeNode tpe, Node<E, T> node) implements Expr<E, T> {
    @Override
    public String prettyPrint(int indentation) {
      return "Ascription(node="
          + node.expr().prettyPrint(indentation + 2)
          + ", tpe="
          + tpe.preType().toString()
          + ")";
    }
  }

\end{lstlisting}
\caption{Combined excerpts from \texttt{AntlrHyggeNodeVisitor.java} and \texttt{Expr.java} showing the visitor and the definition for the \texttt{Ascription}-expression. }
\label{fig:antlr_node_visitor}
\end{figure}

The creation of \texttt{Node}- and \texttt{Expr}-objects is mostly straight forward even though they are mutually recursive.
To parse the argument for the assertion, one simply calls the \texttt{accept()}-method on the subexpression for the visitor
context, which will recursively traverse and transform the AST to \texttt{Node}-instances. Most of the expression types
follow a similar implementation style as the one for assertions. There are some of these tranformations that are more complicated,
like function declarations, structs, unions, etc., but it is the same case for declaring the equivalent parser for \texttt{hyggec}
using \texttt{FsLexYacc}.

To compare the two parser implementations in a more quantitative way, the parser in \texttt{JHygge}
takes up 159 lines (\texttt{hygge.g4}) and 440 lines (\texttt{AntlrHyggeNodeVisitor}) for a total of 599 lines, where the lexer and
parser for \texttt{hyggec} is 100 lines (\texttt{Lexer.fsl}) and 305 lines (\texttt{Parser.fsy}) totalling 405 lines. The Java
implementation takes up approx. $48\%$ more lines than the \texttt{F\#} version, where the \texttt{AntlrHyggeNodeVisitor} is
the largest contributor.

\section{Generating JVM bytecode with the Class-file API from Java 24}

The Class-file API promises a declarative approach to reading, modifying and writing \texttt{.class}-files. The main parts of the
API used in the implementation of the JVM backend in \texttt{JHygge} is the \texttt{ClassBuilder} and the \texttt{CodeBuilder}.
A \texttt{ClassBuilder} is intended for modifying classes at a higher-level, such as adding attributes, constructors and methods,
without directly handling bytecode. A \texttt{CodeBuilder} is then used to construct the sequence of bytecode instructions in
a method or a constructor. Both the APIs for \texttt{ClassBuilder}s and \texttt{CodeBuilder}s make use of the builder pattern,
however they are different in the respect that a \texttt{CodeBuilder} is immutable whereas a \texttt{ClassBuilder} isn't.
We utilize this difference in mutability during the code generation of functions and structs.

Let's walk through an example to show, how the Class-file API is used concretely in \texttt{JHygge}. The code listing  shows
the implementation of the \texttt{generateCode()}-method for the \texttt{JVMCodeGenerator}-class. This is the minimal requirement
for implementing the \texttt{HyggeCodeGenerator}-interface. As we see in the listing, we initially generate an empty \texttt{.class}-file,
which is always \texttt{HyggeMain.class}, containing the \texttt{HyggeMain} class. The \texttt{ClassBuilder}-API is then used to
generate the static \texttt{main()}-method, which the JVM will look for and use as an entry-point. To generate the code for
the body of the \texttt{main()}-method, we use the immutable \texttt{CodeBuilder}-API. Similar to the typechecker and the interpreter,
code generation is split across different classes and methods by grouping language features together, so the overloaded and private
\texttt{generateCode()}-method will then dispatch to the correct code generator method depending on the top-level node in the AST.
Note also, that the \texttt{JVMCodeGeneratorEnvironment}-class keeps a reference to the parent \texttt{ClassBuilder} for the
\texttt{HyggeMain}-class. This will come in handy later, when generating code for functions, for example.

\begin{figure}[H]
\centering 
\begin{lstlisting}[language=Java]
  // From JVMCodeGenerator.java
  @Override
  public void generateCode(Node<E, T> ast, String fileName) throws IOException {
    var filePath = Path.of("HyggeMain.class").toAbsolutePath();
    System.out.println("Compiling '" + fileName + "' to: " + filePath);
    ClassFile.of()
        .buildTo(
            filePath,
            ClassDesc.of("HyggeMain"),
            classBuilder -> {
              var codeGeneratorEnvironment = new JVMCodeGeneratorEnvironment(classBuilder);
              classBuilder.withMethodBody(
                  "main",
                  ofDescriptor("([Ljava/lang/String;)V"),
                  ClassFile.ACC_PUBLIC | ClassFile.ACC_STATIC,
                  codeBuilder ->
                      generateCode(ast, codeGeneratorEnvironment, codeBuilder).return_());
            });
  }
\end{lstlisting}
\caption{}
\label{}
\end{figure}

Depending on the nodes in the AST, we then end up in the appropriate case. For a simple case, let us consider the case
for value-expressions for the basic built-in types of Hygge, \texttt{UnitVal}, \texttt{BoolVal}, etc. For the \texttt{UnitVal}-case,
the simplest thing is to do nothing and return an unmodified code builder. The rest of the value expressions all follow a similar approach,
loading a constant onto the stack with the \texttt{ldc} instruction, except for the \texttt{BoolVal} expression which makes use of the
\texttt{iconst} instructions instead. With the \texttt{ldc()}-method, the Class-file API places the value in the run-time constant pool
of the \texttt{HyggeMain} class and automatically inserts the correct reference to the constant in the pool. This is one of the ways
that the Class-file API provides abstraction while still giving the programmer control over exactly which instructions are generated.
In figure \ref{fig:jvm_value_codegen}, we show the
simple implementation of JVM bytecode generation for Hygge values.

\begin{figure}[H]
\centering
\begin{lstlisting}[language=Java]
  // From JVMCodeGenerator.java
  public CodeBuilder generateCode(
      Node<E, T> ast, JVMCodeGeneratorEnvironment env, CodeBuilder codeBuilder) {
    return switch (ast.expr()) {
      case UnitVal<E, T> _ -> codeBuilder;
      case BoolVal<E, T> v -> v.value() ? codeBuilder.iconst_1() : codeBuilder.iconst_0();
      case IntVal<E, T> v -> codeBuilder.ldc(v.value());
      case FloatVal<E, T> v -> codeBuilder.ldc(v.value());
      case StringVal<E, T> v -> codeBuilder.ldc(v.value());
      // Remaining cases ... 
    }
  }
 
\end{lstlisting}
\caption{An excerpt from \texttt{JVMCodeGenerator.java} for the simple case of generating JVM bytecode for Hygge values using the Class-file API.}
\label{fig:jvm_value_codegen}
\end{figure}

A more intriguing example would be the code generation for \texttt{if}-expressions.

\begin{figure}[H]
\centering 
\begin{lstlisting}[language=Java]
// From JVMControlFlowCodeGenerator.java
public class JVMControlFlowCodeGenerator<E extends TypingEnvironment, T extends Type> {

  public CodeBuilder generateCode(
      JVMCodeGenerator<E, T> codeGenerator,
      JVMCodeGeneratorEnvironment env,
      CodeBuilder codeBuilder,
      Node<E, T> node,
      If<E, T> ifExpr) {
    Label labelTrue = codeBuilder.newLabel();
    Label labelFalse = codeBuilder.newLabel();

    // Generate code for the condition expression and jump to true branch
    // if condition is not 0
    codeBuilder = codeGenerator.generateCode(ifExpr.condition(), env, codeBuilder).ifne(labelTrue);

    // Generate code for false branch
    codeBuilder =
        codeGenerator
            .generateCode(ifExpr.ifFalse(), env, codeBuilder)
            .goto_(labelFalse)
            .labelBinding(labelTrue);

    // Generate code for true branch
    return codeGenerator.generateCode(ifExpr.ifTrue(), env, codeBuilder).labelBinding(labelFalse);
  }
 
  // Remaining methods ...
}
\end{lstlisting}
\caption{An excerpt from \texttt{JVMControlFlowCodeGenerator.java} showing code generation for Hygge \texttt{if}-expressions}
\label{fig:jvm_if_expressions}
\end{figure}

Generating JVM bytecode for functions... In figure \ref{fig:jvm_function_builder}, the \texttt{generateCode()}-method is shown.

\begin{figure}[H]
 \centering 
 \begin{lstlisting}[language=Java]
  public CodeBuilder generateCode(
      JVMCodeGenerator<E, T> codeGenerator,
      JVMCodeGeneratorEnvironment env,
      CodeBuilder codeBuilder,
      Node<E, T> node,
      Lambda<E, T> lambdaExpr) {
    var methodEnv = env.with(new HashMap<>());
    for (var arg : lambdaExpr.args()) {
      methodEnv.store(
          arg.name(),
          pretypeResolver.resolvePretype(methodEnv.getTypingEnvironment(), arg.preType()));
    }
    env.getParentClass()
        .withMethodBody(
            env.getNameStack().peek(),
            ofDescriptor(
                JVMCodeGenerationUtility.makeDescriptor(
                    lambdaExpr.args().stream()
                        .map(
                            arg ->
                                pretypeResolver.resolvePretype(
                                    env.getTypingEnvironment(), arg.preType()))
                        .toList(),
                    lambdaExpr.body().type(),
                    env)),
            ClassFile.ACC_PUBLIC | ClassFile.ACC_STATIC,
            methodCodeBuilder ->
                JVMCodeGenerationUtility.inferReturn(
                    codeGenerator.generateCode(lambdaExpr.body(), methodEnv, methodCodeBuilder),
                    lambdaExpr.body().type(),
                    env));
    return codeBuilder;
  }
 \end{lstlisting}
 \caption{Excerpt of the implementation of the \texttt{generateCode()}-method handling Hygge \texttt{Lambda}-expressions in \texttt{JVMFunctionCodeGenerator.java}.}
 \label{fig:jvm_function_builder}
\end{figure}

For the final example, we showcase the implementation of code generation for Hygge \texttt{struct}s in figure \ref{fig:jvm_struct_builder}.

\begin{figure}[H]
  \centering
  \begin{lstlisting}[language=Java]
    
  private CodeBuilder generateConstructorMethodBody(
      CodeBuilder constructorBuilder,
      JVMCodeGeneratorEnvironment env,
      StructCons<E, T> structConsExpr,
      String structClassName) {
    constructorBuilder =
        constructorBuilder
            .aload(0)
            .invokespecial(
                ClassDesc.ofInternalName("java/lang/Object"),
                "<init>",
                MethodTypeDesc.ofDescriptor("()V"));

    var i = new AtomicInteger(1);
    for (var field : structConsExpr.fields().entrySet()) {
      var fieldType = field.getValue().type();
      var fieldEnv = field.getValue().env();
      var slot = i.getAndIncrement();
      constructorBuilder =
          loadValueOfType(constructorBuilder.aload(0), fieldEnv, fieldType, slot)
              .putfield(
                  ClassDesc.of(structClassName),
                  field.getKey(),
                  ClassDesc.ofDescriptor(
                      JVMCodeGenerationUtility.makeDescriptor(field.getValue().type(), env)));
    }

    return constructorBuilder.return_();
  }

  public CodeBuilder generateCode(
      JVMCodeGenerator<E, T> codeGenerator,
      JVMCodeGeneratorEnvironment env,
      CodeBuilder codeBuilder,
      Node<E, T> node,
      StructCons<E, T> structConsExpr) {
    var structClassName = JVMCodeGenerationUtility.getTypeName(node.type(), node.env());
    var structClassDescriptor = ClassDesc.of(structClassName);
    var constructorDescriptor =
        MethodTypeDesc.ofDescriptor(
            "("
                + structConsExpr.fields().values().stream()
                    .map(value -> JVMCodeGenerationUtility.makeDescriptor(value.type(), env))
                    .collect(Collectors.joining())
                + ")V");

    var filePath = Path.of(structClassName + ".class").toAbsolutePath();
    try {
      ClassFile.of()
          .buildTo(
              filePath,
              structClassDescriptor,
              classBuilder -> {
                classBuilder.withFlags(ClassFile.ACC_PUBLIC);
                for (var field : structConsExpr.fields().entrySet()) {
                  classBuilder.withField(
                      field.getKey(),
                      ClassDesc.ofDescriptor(
                          JVMCodeGenerationUtility.makeDescriptor(field.getValue().type(), env)),
                      ClassFile.ACC_PUBLIC);
                }
                classBuilder.withMethodBody(
                    "<init>",
                    constructorDescriptor,
                    ClassFile.ACC_PUBLIC,
                    constructorBuilder ->
                        generateConstructorMethodBody(
                            constructorBuilder, env, structConsExpr, structClassName));
              });
    } catch (IOException e) {
      throw new CodeGenerationError(
          "ERROR - Could not generate class file for struct '"
              + structClassName
              + "', caused by: "
              + e);
    }

    codeBuilder = codeBuilder.new_(ClassDesc.of(structClassName)).dup();

    for (var field : structConsExpr.fields().entrySet()) {
      codeBuilder = codeGenerator.generateCode(field.getValue(), env, codeBuilder);
    }

    return codeBuilder.invokespecial(structClassDescriptor, "<init>", constructorDescriptor);
  }
  \end{lstlisting}
  \caption{Excerpt of the implementation of code generation for Hygge \texttt{structs}s from \texttt{JVMStructCodeGenerator.java}.}
  \label{fig:jvm_struct_builder}
\end{figure}
